---
title: "Getting started with odumesl"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with odumesl}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 8, fig.height = 3, dpi = 96)
```

**odumesl** provides ggplot2 layers for the seven canonical symbols of
H.T. Odum's Energy Systems Language (ESL) — a visual vocabulary used
across systems ecology and emergy accounting since Odum 1971.

## The seven ESL symbols

```{r, fig.width = 12, fig.height = 2.5}
library(ggplot2)
library(odumesl)

d <- data.frame(
  x    = c(0, 3, 6, 9, 12, 15, 18),
  y    = 0,
  role = c("source", "producer", "consumer", "storage",
           "interaction", "heat_sink", "money"),
  fill = c("#F1C40F", "#228B22", "#C0392B", "#1F77B4",
           "#F39C12", "#7F7F7F", "white"))

ggplot(d, aes(x = x, y = y, fill = I(fill))) +
  geom_odum_source     (data = d[1, ], radius = 0.5) +
  geom_odum_producer   (data = d[2, ], width = 1.5, height = 0.9) +
  geom_odum_consumer   (data = d[3, ], width = 1.5, height = 0.9) +
  geom_odum_storage    (data = d[4, ], width = 1.5, height = 0.9) +
  geom_odum_interaction(data = d[5, ], width = 0.9, height = 0.9) +
  geom_odum_heat_sink  (data = d[6, ], width = 0.6, height = 0.6) +
  geom_odum_money      (data = d[7, ], radius = 0.4,
                        colour = "#B5791B") +
  geom_text(aes(y = -1.5, label = role), size = 3.5) +
  coord_cartesian(xlim = c(-1, 19), ylim = c(-2, 1)) +
  theme_void()
```

## Semantics

Each symbol has a specific meaning in Odum's framework:

- **Source** (circle) — a renewable or external input to the system.
  Sun, wind, tide, hydrocarbon deposits, uranium.
- **Producer** (bullet-nosed hexagon) — an autocatalytic unit that
  captures and concentrates emergy. Plants, agriculture, power
  stations. The pointed nose signals directional output.
- **Consumer** (hexagon) — an entity that uses stored emergy. A
  household, a sector, a cultural producer.
- **Storage** (bullet-tank) — a reservoir of matter, energy, or
  information. Fuel reserves, grid capacity, batteries.
- **Interaction** (diamond) — a multiplicative junction where two or
  more flows combine.
- **Heat sink** (down-triangle) — dispersed (unusable) heat at the
  end of a transformation. Every transformation eventually
  terminates in a heat sink per the second law.
- **Money** (outlined circle, dashed connecting lines) — a
  transaction node, drawn with money flow going in the *opposite*
  direction of energy flow.

## Composing a small diagram

Two producers feeding a storage feeding two consumers with a heat
sink at each stage:

```{r, fig.width = 9, fig.height = 4}
nodes <- data.frame(
  x    = c(0, 0,  3,  6, 6, 3, 6),
  y    = c(1, -1, 0,  1, -1, -2, -2),
  role = c("source1", "source2", "storage",
           "consumer1", "consumer2", "hs1", "hs2"))

ggplot() +
  geom_odum_source     (data = nodes[1:2, ],
                        aes(x = x, y = y), radius = 0.35) +
  geom_odum_storage    (data = nodes[3,   ],
                        aes(x = x, y = y), width = 1.2, height = 0.8) +
  geom_odum_consumer   (data = nodes[4:5, ],
                        aes(x = x, y = y), width = 0.9, height = 0.6) +
  geom_odum_heat_sink  (data = nodes[6:7, ],
                        aes(x = x, y = y), width = 0.4, height = 0.4) +
  # Energy flows
  geom_segment(aes(x = 0.3, y =  1, xend = 2.5, yend = 0.15),
               arrow = arrow(length = unit(0.1, "in"))) +
  geom_segment(aes(x = 0.3, y = -1, xend = 2.5, yend = -0.15),
               arrow = arrow(length = unit(0.1, "in"))) +
  geom_segment(aes(x = 3.7, y = 0.2, xend = 5.5, yend =  1),
               arrow = arrow(length = unit(0.1, "in"))) +
  geom_segment(aes(x = 3.7, y = -0.2, xend = 5.5, yend = -1),
               arrow = arrow(length = unit(0.1, "in"))) +
  coord_cartesian(xlim = c(-0.5, 7), ylim = c(-3, 2)) +
  theme_void()
```

## Low-level helpers

If you need the polygon coordinates directly (for `gganimate`
`transition_*()`, patchwork composition, non-ggplot renderers), the
seven `odum_*` vertex-tibble helpers are also exported:

```{r}
odum_producer(cx = 0, cy = 0, w = 2, h = 1, id = "p1")
```

Each returns a tibble of `(x, y, id)` suitable for `geom_polygon()`.

## Prior art

To the author's knowledge, `odumesl` is the first R (or Python)
grammar-of-graphics library for ESL. The closest peers are the
now-abandoned Java `EmSim` (Valyi & Ortega 2004, an ODE integrator
with no rendering API), the TypeScript GUI editor
[sholtomaud/odum-energy-language](https://github.com/sholtomaud/odum-energy-language),
and static SVG references from the University of Florida's
[Center for Environmental Policy](https://cep.ees.ufl.edu/emergy/resources/symbols_diagrams.shtml)
and the [International Society for the Advancement of Emergy
Research](http://www.emergysociety.com/esl-symbols/).

## References

- Odum, H.T. (1971). *Environment, Power, and Society*. Wiley-Interscience.
- Odum, H.T. (1994). *Ecological and General Systems: An Introduction
  to Systems Ecology*. University Press of Colorado.
- Brown, M.T. (2004). "A picture is worth a thousand words: Energy
  systems language and simulation." *Ecological Modelling* 178: 83-100.
