---
title: "Step-by-step: building a food web one geom at a time"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Step-by-step: building a food web one geom at a time}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.align = "center", out.width = "100%",
  fig.width = 9, fig.height = 3.5, dpi = 110,
  dev.args = list(bg = "#f3f8f3"))
```

This vignette watches a food web assemble one `geom_odum_*` layer at
a time. Same data, six versions of the plot — each step adds one
symbol. Copy the final block, tweak the data, and you have your own
diagram.

```{r libs}
library(ggplot2)
library(energese)

nodes <- data.frame(
  x    = c(1, 3.5, 6, 8.5),
  y    = 2,
  role = c("Sun", "Grass", "Grazer", "Predator"))

ink <- "#1d2a1d"
```

## 1. An empty canvas

Start with `ggplot()` on the node data, `coord_fixed()` for
equal-scale drawing, and `theme_void()` for a clean background.

```{r step-1}
base <- ggplot(nodes, aes(x, y)) +
  coord_fixed(xlim = c(0, 10), ylim = c(0, 4)) +
  theme_void()
base
```

## 2. Add the sun (source)

`geom_odum_source()` — a filled circle. `radius` controls the size.

```{r step-2}
p2 <- base +
  geom_odum_source(data = nodes[1, ], radius = 0.35, fill = "#F1C40F")
p2
```

## 3. Add the primary producer (grass)

`geom_odum_producer()` — the Odum "bullet" shape. Autocatalytic units
that capture and concentrate emergy.

```{r step-3}
p3 <- p2 +
  geom_odum_producer(data = nodes[2, ], width = 1.4, height = 1.0,
                      fill = "#2E7D32")
p3
```

## 4. Draw the sun→grass flow

Arrows between symbols are just `annotate("segment", ...)` calls with
an `arrow()` argument — no special geom needed.

```{r step-4}
p4 <- p3 +
  annotate("segment", x = 1.4, xend = 2.8, y = 2, yend = 2,
           arrow = arrow(length = unit(0.12, "cm")), colour = ink)
p4
```

## 5. Add the grazer and its flow

Grazers are `geom_odum_consumer()` (a hexagon). Then draw the
grass→grazer arrow.

```{r step-5}
p5 <- p4 +
  geom_odum_consumer(data = nodes[3, ], width = 1.3, height = 0.9,
                      fill = "#8B4513") +
  annotate("segment", x = 4.2, xend = 5.35, y = 2, yend = 2,
           arrow = arrow(length = unit(0.12, "cm")), colour = ink)
p5
```

## 6. Add the top predator + its flow

Same geom family, different fill. Grazer→predator arrow closes the
chain.

```{r step-6}
p6 <- p5 +
  geom_odum_consumer(data = nodes[4, ], width = 1.3, height = 0.9,
                      fill = "#5D4037") +
  annotate("segment", x = 6.65, xend = 7.85, y = 2, yend = 2,
           arrow = arrow(length = unit(0.12, "cm")), colour = ink)
p6
```

## 7. Second-law bookkeeping: heat sinks

Every transformation dissipates heat. Add one `geom_odum_heat_sink()`
under each consumer.

```{r step-7}
sinks <- data.frame(x = c(3.5, 6, 8.5), y = 0.7)

p7 <- p6 +
  geom_odum_heat_sink(data = sinks, width = 0.55, height = 0.85,
                       fill = "#455a64") +
  annotate("segment", x = 3.5, xend = 3.5, y = 1.5, yend = 1.1,
           arrow = arrow(length = unit(0.12, "cm")), colour = ink) +
  annotate("segment", x = 6, xend = 6, y = 1.55, yend = 1.1,
           arrow = arrow(length = unit(0.12, "cm")), colour = ink) +
  annotate("segment", x = 8.5, xend = 8.5, y = 1.55, yend = 1.1,
           arrow = arrow(length = unit(0.12, "cm")), colour = ink)
p7
```

## 8. Labels + final polish

Label each node, tighten the frame, add a title. Now it's a diagram.

```{r step-8}
p8 <- p7 +
  geom_text(aes(y = 3.15, label = role), size = 3.2, colour = ink) +
  labs(title = "Solar-driven grazing food web") +
  theme(plot.title = element_text(hjust = 0.5, size = 12, colour = "#176b17"))
p8
```

## What just happened

Every ESL diagram is the same recipe: an `x/y` node table, one
`geom_odum_*` per node kind, `annotate("segment", ...)` for the
flows, `geom_odum_heat_sink()` under every transformation, labels
via `geom_text()`. That's the entire grammar. See the case-studies
vignette (`vignette("case-studies", package = "energese")`) for four
worked classical Odum systems built on this same recipe.
